World Heritage Sites on Wikipedia: Cultural Heritage Activism in a Context of Constrained Agency

### 2021-03-31 analysis/paper/paper.Rmd
many small edits and change csl

---
title: "World Heritage Sites on Wikipedia: Cultural Heritage Activism in a Context of Constrained Agency"
author:
  - Ben Marwick
  - Prema Smith
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::word_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  UNESCO World Heritage sites are places of outstanding significance, and often key sources of information that influence how people interact with the past today. The process of inscription on the UNESCO list is complicated and intersects with political and commercial controversies. But how well are these controversies known to the public? Wikipedia pages on these sites offer a unique dataset for insights into public understanding of heritage controversies. The unique technicity of Wikipedia, with its bot ecosystem and editing mechanics, shapes how knowledge about cultural heritage is constructed, and how controversies are negotiated and communicated. In this article we investigate the patterns of production, consumption and spatial and temporal distributions of Wikipedia pages for World Heritage cultural sites. We find that Wikipedia provides a distinctive context for investigating how people experience and relate to the past in the present. The agency of participants is highly constrained, but distinctive, behind-the-scenes expressions of cultural heritage activism are evident. Concerns about state-like actors, violence and destruction, deal-making, etc. in the World Heritage inscription process are present, but rare on Wikipedia’s World Heritage pages. Instead hyper-local, and process issues dominate controversies on Wikipedia. We describe how this kind of research, drawing on big data and data science methods, contributes to digital heritage studies, and also reveals its limitations.
---

@@

While social media is a vast and diverse online space that we are only just beginning to explore how people engage with the past, there are other contexts of online interactions where heritage is practiced in distinctive, if poorly understood, ways. We can contrast social media, with its fundamental elements of identity, conversations, sharing, presence, relationships, reputation, and groups [@kietzmann2011social], with online peer production communities, where users participate in the collaborative, asynchronous creation, sharing, promoting, and classification of content in highly structured and goal-directed ways [@wilkinson2008strong]. Online peer production communities are comparable to more traditions kinds of voluntary associations where groups set and execute goals, with explicitly democratic organizational ideals. While the ideals of many online peer production communities emphasize non-hierarchical and non-bureaucratic organization, analysis of large amounts of user activity indicates that most of these communities are actually undemocratic and noninclusive, functioning as entrenched oligarchies [@shaw2014laboratories]. This emphasis on governance and management of collective action is a key detail that distinguishes social media from online peer production. It follows that user interactions in the process of generating content in online peer production communities includes technological and social mechanisms that enact the community's governance policies, for example, limiting a user's activity according to their status in the community's hierarchy, or managing conflict with highly structured procedures. Here we show how the distinctive organizational and technical qualities of online peer production communities make them a unique context of heritage production to study digital traces of human activity resulting from engagement with the past.

@@

Here we present a study of how people engage with elements of the past in one of the largest and long-lived online peer production communities, the English-language Wikipedia. Originating in 2001, this is a highly influential and well-known online encyclopedia, that anyone can edit, currently with 11 billion page views per month (<https://stats.wikimedia.org/#/en.wikipedia.org>). Although anyone can edit, most internet users do not. Factors that strongly predict if a user has ever edited Wikipedia include their gender (male), age (younger), education level (has BA), Internet use frequency (higher) and Internet use skills (higher) [@shaw2018pipeline; @hill2013wikipedia; @adams2015wikipedia; @ford2017anyone]. There are also geographical disparities. Articles about rural areas have systematically lower quality, are less likely to have been produced by contributors who focus on the local area, and are more likely to have been generated by bots (automated software agents) [@johnson2016not]. These studies indicate that participation in online peer production communities often follow existing patterns of social exclusion. @graham2014uneven examined the global distribution of 3.4 million geotagged Wikipedia articles and find a pattern of places in the Global North being represented in local languages, while articles about places in the Global South are largely being written by others.

An additional consideration for understanding participation in online peer production communities are the technical schemas of the software that Wikipedia runs on, MediaWiki. This is a complex toolkit that enables participation in Wikipedia in highly structured ways. On one hand, these structured behaviours produce structured datasets that are well suited to data science methods for efficient computational analysis of large numbers Wikipedia articles. On the other hand, they constrain and limit the agency of the user, canalizing their behaviour into a small number of possible actions and acceptable modes of discourse and engagement with other users [@iba2010analyzing]. While Wikipedia has elements that are ubiquitous on the Internet, such as links on articles that take the user to other articles or other pages on the Internet, it also has several less common elements that contribute to its unique technicity, resulting in specific types of relationships between human users and the technical elements of the Wikipedia project [@niederer2010wisdom; @weltevrede2016platform]. For example, every edit to an article is tracked in a publicly accessible version control system attached to that article. This exposes the article creation process in highly granular detail; for any given article, we can see how many editors contributed, the size of their edits and their distribution over time, among other things [@priedhorsky2007creating]. Wikipedia has a special category of edit called the 'revert' which allows a user to restore an article to an earlier state to remove recent vandalism (such as the addition of irrelevant or offensive material). This special revert action, combined with a 'talk' page attached to each encyclopedia article for threaded discussion among editors, allows us to detect and study the social dynamics arising from the creation and editing of articles, for example the controversiality of an article [@yasseri2012dynamics; @suh2007us]. While articles themselves must be written to conform to the fundamental Wikipedia policy of NPOV (Neutral Point Of View), the talk page is where different views are expressed and negotiated among editors.

@@

We investigate how the unique technicity of Wikipedia shapes interactions between people and the past with a case study of cultural sites inscribed on the WHL. We chose the CS-WHL as a bounded set of cultural heritage elements with several characteristics that make it of general interest. It has a global geographic distribution, broad public interest at local and international scales, in both online and face-to-face communities; a wide temporal distribution in both the age of the cultural sites, the ages of inscription on the WHL, and the ages of their appearance on Wikipedia; and finally many CS-WHL have a high intensity of cultural and political discussions that surround events affecting these sites, such as their inscription on the WHL. These qualities make it an ideal data set as an entry point for case studies of digital heritage in online peer production communities, where activities are typically goal-driven (e.g. 'write quality articles') compared to social media activity where user activities are more often event-driven (e.g. 'share reactions to Brexit').

@@

Physical conflicts at or near CS-WHL are major events that also galvanize public interest in these locations. World Heritage sites in Palestine, Mali, Syria, Congo and Cambodia have recently been sites of violence, in many cases specifically linked to their potential WHL nomination, listing, or management. In 1998 anti-government and mostly Hindu Tamil groups bombed the holy Buddhist site of the Temple of the Tooth at the WHL site of Kandy (the last capital of the ancient kings of Sri Lanka), killing 17 people and substantially damaging the temple [@coningham1999paradise]. In Mali during 2012, fighting between government and rebel groups lead to the damage and destruction of tombs at the CS-WHL sites of Gao and Timbuktu [@brioschi2017discourses]. The World Heritage Committee found itself powerless to intervene because of political gridlock [@meskell2015gridlock], and these Mali sites are currently among the 36 cultural sites on the List of World Heritage in Danger (<https://whc.unesco.org/en/danger/>). In 2015, ISIS militants destroyed the Temple of Bel in Palmyra, Syria (a CS-WHL site of monumental ruins, once great city at the crossroads between east and west in the ancient world) [@gornik2015can]. Preah Vihear, inscribed in 2008, is a CS-WHL located on a long-disputed section of the Thai-Cambodia border that has been a site of both violent military clashes and international political intrigue. Although both Thailand and Cambodia supported the nomination of the site to the WHL, the Thai government objected to maps in the nomination package that showed Cambodia as the owner of disputed land next to the temple, leading to protests and military clashes [@sothirak2013cambodia]. US diplomatic cables released by WikiLeaks reveal that settlement of disputes over Preah Vihear were intricately tied to broader issues of foreign policy and US and Chinese investment, especially access to natural gas reserves in the Gulf of Thailand [@meskell2020world].

@@

Our brief review of controversial cultural sites on the WHL shows the intensity and diversity of conflicts and tensions that surround these sites. Many CS-WHL are symbols of national, cultural, political, and religious identity, and the extent of political involvement in negotiations of WHL inscriptions indicates they are of great public interest among local and diasporic communities. Our goal in this study is to answer the question of how this interest is expressed within the socio-technical constraints of the English-language Wikipedia. We surveyed the basic characteristics of content (article length, number of wikilinks out to other pages, number of citations to non-Wikipedia items), consumption (page view counts, wikilinks in from other Wikipedia pages), and production (edit counts, edit densities, edit sizes, number of unique editors per article, talk page length, talk page topics). By comparing these basic characteristics of English-language Wikipedia articles about CS-WHL to 10,000 random English-language Wikipedia articles we can approach the question: can metrics of content, consumption, and production indicate engagement with the past via CS-WHL on Wikipedia? Can we detect conflict in the edit histories, bot activity, and talk pages for Wikipedia articles about CS-WHL sites, and how does this conflict relate to the types of controversies noted above? Random articles were obtained by sending GET requests to the 'random' module in the Wikimedia REST API (<https://en.wikipedia.org/api/rest_v1/>).

The highly detailed edit histories that Wikipedia keeps for every article allow us to further investigate spatial and temporal questions relating to engagement with the past and conflicts surrounding CS-WHL sites. For example, when an article is anonymously edited, for example, by a user who does not have a Wikipedia user account (or is not logged into their account), their edit is identified by that person's IP address. Edits made by people who are logged in to their Wikipedia user account do not include the user's IP address, only their Wikipedia user account name. An IP address can be used to geo-locate the user to the country they were in when the made the edit. This means that edits from registered Wikipedia users cannot be used for tracing the geographic origin of an edit, but anonymous edits, because they are identified by an IP address, can be used for tracing the geographic origin of an edit. We geo-located all edits with IP addresses for all English-language Wikipedia articles CS-WHL sites to determine the country of origin of those edits using the rgeolocate package for R [@rgeolocate]. This helps us to answer the question: are the editors of articles about CS-WHL located near the sites they edit, indicating local community interest in the online representation of their heritage? The time and date stamps attached to every edit on every article allows us to investigate temporal patterns of activity on CS-WHL Wikipedia articles. Analyses of these data help us to answer the question: is Wikipedia editing activity correlated with events outside of Wikipedia relating to the CS-WHL sites, such as conflict events, or their inscription on the WHL?

@@

Of the `r nrow(wh_unesco)` cultural sites on the World Heritage List at the time of writing, we found Wikipedia articles for `r nrow(some_page_variables)`. As a group, the basic details of content for Wikipedia articles for CS-WHL differ little from a sample of 10,000 random Wikipedia articles (Figure \@ref(fig:wh-wiki-content)). The scholarly nature of the articles, measured by the number of sources cited in the reference list per thousand words in the article body, have similarly distributions for CS-WHL articles and random articles. The number of Wikilinks out from the target article to other Wikipedia articles are also similarly distributed for CS-WHL articles and random articles. The total number of words in a CS-WHL article is typically much higher than a random article, indicating that they receive more generative effort from editors than other articles.

@@

Although details of content of CS-WHL articles are similar to our random sample, variables related to the production of Wikipedia articles on CS-WHL differ in important ways from other articles (Figure \@ref(fig:wh-wiki-production)). The number of edits per thousand words, or edit density, and the number of unique editors per thousand words, or editor density, are substantially higher for CS-WHL articles. This tells us that CS-WHL articles are intensively word-smithed by a more diverse community of editors than for other articles. The absolute size of edits (i.e. additions or removals of text) is about the same for CS-WHL articles as other articles. The involvement of bots in producing CS-WHL articles is also about the same as for other articles. Bot activity is most intense on shorter, low-profile CS-WHL articles, in Figure \@ref(fig:wh-wiki-production) the labeled points are sites where bots have done \>30% of edits. The most active bot on CS-WHL articles is Cluebot NG (vandalism detection and reverting), compared to Cydebot (automatic implementation of category deletions) for the random articles. The AnomieBOT, which performs clerical duties in an article's reference list, is highly active on CS-WHL articles compared to random articles, reflecting the longer reference lists that CS-WHL articles tend to have relative to other articles. Most bot edits on on CS-WHL articles are in the fixer, tagger, connector, and clerk roles [@zheng2019roles]. None of these articles with intensive bot activity are CS-WHL sites of conflict or on the List of World Heritage in Danger, indicating that these sites receive little or no vandalism.

@@

The talk page for Mapungubwe, which is longer than the article itself, is lengthy with expressions of concern about the accuracy of information in the article. In particular, editors were concerned about the depiction of Indigenous people, especially the degree to which early European colonists were aware of Indigenous communities, and the correct cultural affiliation of the Indigenous groups that originally occupied the site. In this remarkable case we see the Wikipedia editors engaging conflicts that go beyond the typical technical details of article production. What is especially notable about these discussions on the talk page is how closely they echo the debates of land ownership that complicated the WHC process [@meskell2011paris]. The editors involved in this debate did not make any edits to the article itself which raises questions about their motivations for engaging in debate on the talk page, since they do not seem personally invested in the content of the article. Without directly interviewing these editors, we cannot be sure of their motives for participating in the talk page.

A second important example of conflict on a talk page is the article about Preah Vihear Temple. Although not notable for the length of its talk page, the discussions among editors of the article about Preah Vihear Temple include a strongly worded disagreement among seven editors about the legitimacy of the ownership claims by Thailand and Cambodia of the territory that includes the temple. As for Mapungubwe, the debate among Wikipedia editors mirrors closely the political tensions surrounding the nomination of Preah Vihear. The talk page also has a lengthy argument from to 2008, involving several editors, about whether or not the Thai name of the site should be included in the text of the article. The editors appear to be people of Thai and Khmer heritage, with comments that personalise the national tensions such as "we invaded you", and references to contemporary political tensions such as the 2008 street demonstrations in Thailand that saw conflict between the royalist People's Alliance for Democracy and the populist People's Power Party. One editor indicates their likely Thai ethnicity here through their use of Thai language phrases (using the Thai alphabet). This debate also recapitulates broader substantive issues of whether Thailand or Cambodia country has the stronger claim for ownership of the temple. Amid accusations of nationalism among the editors, one of them asks "Can we put aside politics and be more collaborative on the encyclopedia project?" and appeals to Wikipedia policies 'Third opinion' and 'Dispute resolution' to diffuse the tensions and refocus attention on the common goal. Unlike the Mapungubwe talk page, many of the editors involved in the Preah Vihear talk page debates have made contributions directly to the Preah Vihear article, and are deeply invested in how the article represents the site.

@@

The distribution of editor locations shown in figure \@ref(fig:wh-wiki-circle) was determined by geo-locating the IP addresses attached to each anonymous edit on all CS-WHL articles (only anonymous edits contain IP addresses, edits made by registered Wikipedia users contain no information suitable for geolocation). This figure shows the flow of edits, with the arrows starting at the country where the editor was located when they made their edit, and ending at the country where the CS-WHL site that they are editing is located. The inset plot on figure \@ref(fig:wh-wiki-circle) shows that this visualisation only accounts for a relatively small proportion of all edits to CS-WHL articles. Around 20% of edits of most CS-WHL articles are anonymous, a higher proportion than our random sample. Nevertheless, there are `r prettyNum(nrow(revision_history_page_details_ip_addresses_df), big.mark=",")` anonymous edits, and so this sample likely has some informational value in characterizing the spatial distribution of edits. The most striking detail in figure \@ref(fig:wh-wiki-circle) is the large proportion of edits that originate in the United States, and that the vast majority of these are edits of articles about CS-WHL sites located elsewhere in the world. The country with the next largest proportion, the United Kingdom, is less than half of the United States, and a much greater proportion of edits originating in the UK are on articles about CS-WHL sites located in the UK, compared to the US. The country with the next highest proportion of editors is India, and nearly half of those edits are on CS-WHL sites located India. Generally, articles about CS-WHL sites receive edits from editors located elsewhere, mostly the US and other English-speaking countries.

@@

Time series analysis of edits on articles about CS-WHL sites can help us identify relationships between activity on Wikipedia articles and external events related to the sites. Figure \@ref(fig:wh-wiki-time1) shows that most articles about CS-WHL sites are much older than articles in our random sample, with most CS-WHL articles created 5-10 years after Wikipedia first appeared in 2001. This rapid addition of articles early in the life of Wikipedia supports the observation from the in-bound Wikilinks data that articles about CS-WHL sites are more central to the encyclopedia and considered worthy of inclusion than many other types of articles. The distribution of years between the date of a site's inscription on the WHL and the date of the creation of that site's Wikipedia article has a distinctive trimodal shape. The first two modes reflect the relatively large number of sites inscribed on the WHL in 1983 and the early 1990s, before Wikipedia was created in 2001. The third mode at the zero point on the horizontal axis corresponds to CS-WHL sites that were inscribed after the creation of Wikipedia, and have Wikipedia articles that were created in or around the same year the CS-WHL site was inscribed on the list. Only a relatively small proportion of CS-WHL sites had Wikipedia articles created about them before they were inscribed on the WHL, because the majority of sites were inscribed before Wikipedia was created.

The time series of editing activity at individual sites are highly variable and show display complex patterns (Figure \@ref(fig:wh-wiki-time2)). Timbuktu shows spikes of editing activity right after the rebel attacks in mid June 2012, with descriptions of the attacks added to the text of the article at that time. However the tallest peak in editing activity on the Timbuktu occurs slightly earlier, in April 2012, those edits concentrated on expanding the sections on local prehistory. A detailed analysis of all the article-specific time series is beyond the scope of this paper, but we can see several patterns indicating important relationships between editorial activities and WHL listing. For example, one pattern is of editing activity spiking sharply at the time of WHL inscription (e.g. Tyre, Masada), a second pattern is editorial activity spiking at the time of WHL inscription and slowly decaying to a baseline level (e.g. Carthage), a third pattern is editorial activity that peaks shortly after the time of WHL inscription (e.g. Troy, Masada), and finally many sites where editorial activity that shows no signal at all at the time of WHL inscription (Figure \@ref(fig:wh-wiki-time2)).

>>>>

---
title: "World Heritage Sites on Wikipedia: Cultural Heritage Activism in a Context of Constrained Agency"
author:
  - Ben Marwick
  - Prema Smith
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::word_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/harvard-cite-them-right.csl" # Insert path for the bib-style
abstract: |
  UNESCO World Heritage sites are places of outstanding significance, and often key sources of information that influence how people interact with the past today. The process of inscription on the UNESCO list is complicated and intersects with political and commercial controversies. But how well are these controversies known to the public? Wikipedia pages on these sites offer a unique dataset for insights into public understanding of heritage controversies. The unique technicity of Wikipedia, with its bot ecosystem and editing mechanics, shapes how knowledge about cultural heritage is constructed, and how controversies are negotiated and communicated. In this article we investigate the patterns of production, consumption and spatial and temporal distributions of Wikipedia pages for World Heritage cultural sites. We find that Wikipedia provides a distinctive context for investigating how people experience and relate to the past in the present. The agency of participants is highly constrained, but distinctive, behind-the-scenes expressions of cultural heritage activism are evident. Concerns about state-like actors, violence and destruction, deal-making, etc. in the World Heritage inscription process are present, but rare on Wikipedia’s World Heritage pages. Instead, hyper-local, and process issues dominate controversies on Wikipedia. We describe how this kind of research, drawing on big data and data science methods, contributes to digital heritage studies, and also reveals its limitations.
---

@@

While social media, such as Facebook and Twitter, is a vast and diverse online space that we are only just beginning to explore how people use to engage with the past, there are other contexts of online interactions where heritage is practiced in distinctive, if poorly understood, ways. We can contrast social media, with its fundamental elements of identity, conversations, sharing, presence, relationships, reputation, and groups [@kietzmann2011social], with online peer production communities, where users participate in the collaborative, asynchronous, creating, sharing, promoting, and classifying of content in highly structured and goal-directed ways [@wilkinson2008strong]. Online peer production communities are comparable to more traditional kinds of voluntary associations where groups set and execute goals, with explicitly democratic organizational ideals. While the ideals of many online peer production communities emphasize non-hierarchical and non-bureaucratic organization, analysis of large amounts of user activity indicates that most of these communities are actually undemocratic and noninclusive, functioning as entrenched oligarchies [@shaw2014laboratories]. This emphasis on governance and management of collective action is a key detail that distinguishes social media from online peer production. It follows that user interactions in the process of generating content in online peer production communities includes technological and social mechanisms that enact the community's written or unwritten governance policies. These may include, for example, limiting a user's activity according to their status in the community's hierarchy, or managing conflict with highly structured procedures. Here we show how the distinctive organizational and technical qualities of online peer production communities make them a unique context of heritage production to study digital traces of human activity resulting from engagement with the past.

@@

We present a study of how people engage with elements of the past in one of the largest and long-lived online peer production communities, the English-language Wikipedia. Originating in 2001, this is a highly influential and well-known online encyclopedia, that anyone can edit, with 9 billion page views per month as of September 2020 (<https://stats.wikimedia.org/#/en.wikipedia.org>). Although anyone can edit, most internet users do not. Factors that strongly predict if a user has ever edited Wikipedia include their gender (male), age (younger), education level (has BA), Internet use frequency (higher) and Internet use skills (higher) [@shaw2018pipeline; @hill2013wikipedia; @adams2015wikipedia; @ford2017anyone]. There are also geographical disparities. Articles about rural areas have systematically lower quality, are less likely to have been produced by contributors who focus on the local area, and are more likely to have been generated by bots (automated software agents) [@johnson2016not]. These studies indicate that participation in online peer production communities often follow existing patterns of social exclusion. @graham2014uneven examined the global distribution of 3.4 million geotagged Wikipedia articles and find a pattern of places in the Global North being represented in local languages, while articles about places in the Global South are largely being written by others.

An additional consideration for understanding participation in online peer production communities are the technical schemas of MediaWiki, the software that Wikipedia runs on. This is a complex toolkit that enables participation in Wikipedia in highly structured ways. On one hand, these structured behaviours produce structured datasets that are well suited to data science methods for efficient computational analysis of large numbers Wikipedia articles. On the other hand, they constrain and limit the agency of the user, canalizing their behaviour into a small number of possible actions and acceptable modes of discourse and engagement with other users [@iba2010analyzing]. While Wikipedia has elements that are ubiquitous on the Internet, such as links that take the user to other articles or pages on the Internet, it also has several less common elements that contribute to its unique technicity, resulting in specific types of relationships between human users and the technical elements of the Wikipedia project [@niederer2010wisdom; @weltevrede2016platform]. For example, every edit to an article is tracked in a publicly accessible version control system associated with that article. This exposes the article creation process in highly granular detail; for any given article, we can see how many editors contributed, the size of their edits and their distribution over time, among other things [@priedhorsky2007creating]. Wikipedia has a special category of edit called the 'revert' which allows a user to restore an article to an earlier state to remove recent vandalism (such as the addition of irrelevant or offensive material). This special revert action, combined with a 'talk' page attached to each encyclopedia article for threaded discussion among editors, allows us to detect and study the social dynamics arising from the creation and editing of articles, for example the controversiality of an article [@yasseri2012dynamics; @suh2007us]. While articles themselves must be written to conform to the fundamental Wikipedia policy of NPOV (Neutral Point Of View), the talk page is where different views are expressed and negotiated among editors.

@@

We investigate how the unique technicity of Wikipedia shapes interactions between people and the past with a case study of cultural sites inscribed on the WHL. We chose the CS-WHL as a bounded set of cultural heritage elements with several characteristics that make it of general interest. It has a global geographic distribution; broad public interest at local and international scales, in both online and face-to-face communities; a wide temporal distribution in both the age of the cultural sites, the ages of inscription on the WHL, and the ages of their appearance on Wikipedia; and finally, many CS-WHL have a high intensity of cultural and political discussions that surround events affecting these sites, such as their inscription on the WHL. These qualities make it an ideal data set as an entry point for case studies of digital heritage in online peer production communities, where activities are typically goal-driven (e.g. 'write quality articles') compared to social media activity where user activities are more often event-driven (e.g. 'share reactions to Brexit').

@@

Physical conflicts at or near CS-WHL are major events that also galvanize public interest in these locations. World Heritage sites in Palestine, Mali, Syria, Congo and Cambodia have recently been sites of violence, in many cases specifically linked to their potential WHL nomination, listing, or management. In 1998 anti-government and mostly Hindu Tamil groups bombed the holy Buddhist site of the Temple of the Tooth at the WHL site of Kandy (the last capital of the ancient kings of Sri Lanka), killing 17 people and substantially damaging the temple [@coningham1999paradise]. In Mali during 2012, fighting between government and rebel groups lead to the damage and destruction of tombs at the CS-WHL sites of Gao and Timbuktu [@brioschi2017discourses]. The World Heritage Committee found itself powerless to intervene because of political gridlock [@meskell2015gridlock], and these Mali sites are among the 53 cultural sites on the List of World Heritage in Danger, as of March 2021 (https://whc.unesco.org/en/danger/). In 2015, ISIS militants destroyed the Temple of Bel in Palmyra, Syria (a CS-WHL site of monumental ruins, once great city at the crossroads between east and west in the ancient world) [@gornik2015can]. Preah Vihear, inscribed in 2008, is a CS-WHL located on a long-disputed section of the Thai-Cambodia border that has been a site of both violent military clashes and international political intrigue. Although both Thailand and Cambodia supported the nomination of the site to the WHL, the Thai government objected to maps in the nomination package that showed Cambodia as the owner of disputed land next to the temple, leading to protests and military clashes [@sothirak2013cambodia]. US diplomatic cables released by WikiLeaks reveal that settlement of disputes over Preah Vihear were intricately tied to broader issues of foreign policy and US and Chinese investment, especially access to natural gas reserves in the Gulf of Thailand [@meskell2020world].

@@

Our brief review of contentious cultural sites on the WHL shows the intensity and diversity of conflicts and tensions that surround these sites. Many CS-WHL are symbols of national, cultural, political, and religious identity, and the extent of political involvement in negotiations of WHL inscriptions indicates they are of great public interest among local and diasporic communities. Our goal in this study is to answer the question of how this interest is expressed within the socio-technical constraints of the English-language Wikipedia. We surveyed the basic characteristics of content (article length, number of wikilinks out to other pages, number of citations to non-Wikipedia items), consumption (page view counts, wikilinks in from other Wikipedia pages), and production (edit counts, edit densities, edit sizes, number of unique editors per article, talk page length, talk page topics). By comparing these basic characteristics of English-language Wikipedia articles about CS-WHL to 10,000 random English-language Wikipedia articles, we can approach the question: can metrics of content, consumption, and production indicate engagement with the past via CS-WHL on Wikipedia? Can we detect conflict in the edit histories, bot activity, and talk pages for Wikipedia articles about CS-WHL sites, and how does this conflict relate to the types of controversies noted above? Random articles were obtained by sending GET requests to the 'random' module in the Wikimedia REST API (<https://en.wikipedia.org/api/rest_v1/>).

The highly detailed edit histories that Wikipedia keeps for every article allow us to further investigate spatial and temporal questions relating to engagement with the past and conflicts surrounding CS-WHL sites. For example, when an article is anonymously edited, for example, by a user who does not have a Wikipedia user account (or is not logged into their account), their edit is identified by that person's IP address. An IP address can be used to geolocate the user to the country they were in when the made the edit. Edits made by people who are logged in to their Wikipedia user account do not include the user's IP address, only their Wikipedia user account name. This means that edits from registered Wikipedia users cannot be used for tracing the geographic origin of an edit, but anonymous edits, because they are identified by an IP address, can be used for tracing the geographic origin of an edit. We used the rgeolocate package for R [@rgeolocate] to geolocate all edits with IP addresses for all English-language Wikipedia articles CS-WHL sites to determine the country of origin of those edits. This helps us to answer the question: are the editors of articles about CS-WHL located near the sites they edit, indicating local community interest in the online representation of their heritage? The time and date stamps attached to every edit on every article allows us to investigate temporal patterns of activity on CS-WHL Wikipedia articles. Analyses of these temporal data help us to answer the question: is Wikipedia editing activity correlated with events outside of Wikipedia relating to the CS-WHL sites, such as conflict events, or their inscription on the WHL?

@@

Of the `r nrow(wh_unesco)` cultural sites on the World Heritage List at the time of writing, we found Wikipedia articles for `r nrow(some_page_variables)`. As a group, the basic details of content for Wikipedia articles for CS-WHL differ little from a sample of 10,000 random Wikipedia articles (Figure \@ref(fig:wh-wiki-content)). The scholarly nature of the articles, measured by the number of sources cited in the reference list per thousand words in the article body, have similar distributions for CS-WHL articles and random articles. The number of Wikilinks out from the target article to other Wikipedia articles are also similarly distributed for CS-WHL articles and random articles. The total number of words in a CS-WHL article is typically much higher than a random article, indicating that they receive more generative effort from editors than other articles.

@@

Although details of content of CS-WHL articles are similar to our random sample, variables related to the production of Wikipedia articles on CS-WHL differ in important ways from other articles (Figure \@ref(fig:wh-wiki-production)). The number of edits per thousand words, or edit density, and the number of unique editors per thousand words, or editor density, are substantially higher for CS-WHL articles. This tells us that CS-WHL articles are intensively word-smithed by a more diverse community of editors than for other articles. The absolute size of edits (i.e. additions or removals of text) is about the same for CS-WHL articles as other articles. The involvement of bots in producing CS-WHL articles is also about the same as for other articles. Bot activity is most intense on shorter, low-profile CS-WHL articles, in Figure \@ref(fig:wh-wiki-production) the labeled points are sites where bots have done \>30% of edits. The most active bot on CS-WHL articles is Cluebot NG (vandalism detection and reverting), compared to Cydebot (automatic implementation of category deletions) for the random articles. The AnomieBOT, which performs clerical duties in an article's reference list, is highly active on CS-WHL articles compared to random articles. Most bot edits on on CS-WHL articles are in the fixer, tagger, connector, and clerk roles [@zheng2019roles]. None of these articles with intensive bot activity are CS-WHL sites of conflict or on the List of World Heritage in Danger, indicating that these sites receive little or no vandalism.

@@

The talk page for Mapungubwe, which is longer than the article itself, is lengthy with expressions of concern about the accuracy of information in the article. In particular, editors were concerned about the depiction of Indigenous people, especially the degree to which early European colonists were aware of Indigenous communities, and the correct cultural affiliation of the Indigenous groups that originally occupied the site. In this remarkable case we see the Wikipedia editors engaging in conflicts that go beyond the typical technical details of article production. What is especially notable about these discussions on the talk page is how closely they echo the debates of land ownership that complicated the WHC process [@meskell2011paris]. The specific editors involved in this debate did not make any edits to the article itself, which raises questions about their motivations for engaging in debate on the talk page, since they do not seem personally invested in the content of the article. Without directly interviewing these editors, we cannot be sure of their motives for participating in the talk page.

A second important example of conflict on a talk page is the article about Preah Vihear Temple. Although not notable for the length of its talk page, the discussions among editors of the article about Preah Vihear Temple include a strongly worded disagreement among seven editors about the legitimacy of the ownership claims by Thailand and Cambodia of the territory that includes the temple. As for Mapungubwe, the debate among Wikipedia editors here mirrors closely the political tensions surrounding the nomination of Preah Vihear. The talk page also has a lengthy argument from to 2008, involving several editors, about whether or not the Thai name of the site should be included in the text of the article. The editors appear to be people of Thai and Khmer heritage, with comments that personalise the national tensions such as "we invaded you", and references to contemporary political tensions such as the 2008 street demonstrations in Thailand that saw conflict between the royalist People's Alliance for Democracy and the populist People's Power Party. One editor indicates their likely Thai ethnicity here through their use of Thai language phrases (using the Thai alphabet). This debate also recapitulates broader substantive issues of whether Thailand or Cambodia has the stronger claim for ownership of the temple. Amid accusations of nationalism among the editors, one of them asks "Can we put aside politics and be more collaborative on the encyclopedia project?" and appeals to Wikipedia's policies of 'Third opinion' and 'Dispute resolution' to diffuse the tensions and refocus attention on the common goal. Unlike the Mapungubwe talk page, many of the editors involved in the Preah Vihear talk page debates have made contributions directly to the Preah Vihear article, and are deeply invested in how the article represents the site.

@@

The distribution of editor locations shown in figure \@ref(fig:wh-wiki-circle) was determined by geo-locating the IP addresses attached to each anonymous edit on all CS-WHL articles (only anonymous edits contain IP addresses, edits made by registered Wikipedia users contain no information suitable for geolocation). This figure shows the flow of edits, with the arrows starting at the country where the editor was located when they made their edit, and ending at the country where the CS-WHL site that they are editing is located. The inset plot on figure \@ref(fig:wh-wiki-circle) shows that this visualisation accounts for a relatively small proportion of all edits to CS-WHL articles. Around 20% of edits of most CS-WHL articles are anonymous, a higher proportion than our random sample. Nevertheless, there are `r prettyNum(nrow(revision_history_page_details_ip_addresses_df), big.mark=",")` anonymous edits, and so this sample likely has some informational value in characterizing the spatial distribution of edits. The most striking detail in figure \@ref(fig:wh-wiki-circle) is the large proportion of edits that originate in the United States, and that the vast majority of these are edits of articles about CS-WHL sites located elsewhere in the world. The country with the next largest proportion, the United Kingdom, is less than half of the United States, and a much greater proportion of edits originating in the UK are on articles about CS-WHL sites located in the UK, compared to the US. The country with the next highest proportion of editors is India, and nearly half of those edits are on CS-WHL sites located India. Generally, articles about CS-WHL sites receive edits from editors located elsewhere, mostly the US and other English-speaking countries.

@@

Time series analysis of edits on articles about CS-WHL sites help us identify relationships between activity on Wikipedia articles and external events related to the sites. Figure \@ref(fig:wh-wiki-time1) shows that most articles about CS-WHL sites are much older than articles in our random sample, with most CS-WHL articles created 5-10 years after Wikipedia first appeared in 2001. This rapid addition of articles early in the life of Wikipedia supports our earlier observation that articles about CS-WHL sites are more central to the encyclopedia and considered more worthy of inclusion than many other types of articles. The distribution of years between the date of a site's inscription on the WHL and the date of the creation of that site's Wikipedia article has a distinctive trimodal shape. The first two modes reflect the relatively large number of sites inscribed on the WHL in 1983 and the early 1990s, before Wikipedia was created in 2001. The third mode at the zero point on the horizontal axis corresponds to CS-WHL sites that were inscribed after the creation of Wikipedia, and have Wikipedia articles that were created in or around the same year the CS-WHL site was inscribed on the list. Only a relatively small proportion of CS-WHL sites had Wikipedia articles created about them before they were inscribed on the WHL, because the majority of sites were inscribed before Wikipedia was created.

The time series of editing activity at individual sites are highly variable and show display complex patterns (Figure \@ref(fig:wh-wiki-time2)). Timbuktu shows spikes of editing activity right after the rebel attacks in mid June 2012, with descriptions of the attacks added to the text of the article at that time. However the tallest peak in editing activity on the Timbuktu occurs slightly earlier, in April 2012, with numerous edits concentrated on expanding the sections on local prehistory. A detailed analysis of all the article-specific time series is beyond the scope of this paper, but we can see several patterns indicating important relationships between editorial activities and WHL listing. For example, one pattern is of editing activity spiking sharply at the time of WHL inscription (e.g. Tyre, Masada), a second pattern is editorial activity spiking at the time of WHL inscription and slowly decaying to a baseline level (e.g. Carthage), a third pattern is editorial activity that peaks shortly after the time of WHL inscription (e.g. Troy, Masada), and finally many sites where editorial activity that shows no signal at all at the time of WHL inscription (Figure \@ref(fig:wh-wiki-time2)).
